---
title: "P8105 Homework 3"
author: "Christine Lucille Kuryla (clk2162)"
date: "2023-10-14"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

# Problem 1

Load data

```{r}
library(p8105.datasets)
data("instacart")
```

Explore and describe dataset.

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations.

```{r}

# Check out the structure of the data set 
head(instacart)

```

This dataset has `r nrow(instacart)` rows and `r ncol(instacart)` columns. The column names are `r colnames(instacart)`. It is in a long format, and the order_id is not distinct. Each row in the datas et represents a particular item purchased and contains corresponding information about it. There are `r instacart %>% select(order_id) %>% n_distinct()` distinct orders, `r instacart %>% select(user_id) %>% n_distinct()` users (so it seems to be one order per user), containing a total of `r instacart %>% select(product_id) %>% n_distinct()` products.

```{r}

# aisles and number of times they occur (number of times there is an order from the aisle)
# arrange with most ordered from aisles at the top
aisles_df <- instacart %>%
  count(aisle) %>% 
  arrange(desc(n))

head(aisles_df)

```


There are `r nrow(aisles_df)` aisles. The aisles most ordered from are:

```{r aisles_most_orders, echo = FALSE}

aisles_most_orders <- head(aisles_df %>% select("aisle"), n=10)

knitr::kable(aisles_most_orders)

```

```{r instacart_aisle_plot_and_tables}

# Plot number of items ordered in top 10000 aisles
aisles_df %>%
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) +
  geom_point() +
  labs(title = "Number of items ordered in each aisle",
       y = "number of items ordered") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

# Three most popular items in three aisles of interest
instacart %>% 
    filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
      group_by(aisle) %>% 
    count(product_name) |> 
    mutate(rank = min_rank(desc(n))) %>% 
    filter(rank < 4) %>% 
    arrange(desc(n)) %>% 
    knitr::kable()
  
# Pink Lady Apples and Coffee Ice Cream order details
instacart %>% 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>% 
  group_by(product_name, order_dow) %>% 
  summarize(mean_hour = mean(order_hour_of_day)) %>% 
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) %>% 
  knitr::kable(digits = 2)

```

# Problem 2

This problem has to do with several things, we are focusing on self reported health over locations and years.

Load data 

```{r q2_load}
library(p8105.datasets)
data("brfss_smart2010")
```

Clean data

```{r q2_clean}

brfss_smart2010_raw <- brfss_smart2010

brfss_smart2010_overall_health <- brfss_smart2010 %>%
  janitor::clean_names() %>%
  rename(state = locationabbr) %>% 
  filter(topic == "Overall Health") %>%
  mutate(response = fct_relevel(response, c("Poor", "Fair", "Good", "Very good", "Excellent")))

#brfss_smart2010_overall_health_excellent_poor <- brfss_smart2010_overall_health %>%
#  filter(response %in% c("Excellent", "Poor"))
  # include only responses from “Excellent” to “Poor” ???????

knitr::kable(head(brfss_smart2010_overall_health, n = 20))

#knitr::kable(head(brfss_smart2010_overall_health_excellent_poor, n = 20))

```

In 2002, which states were observed at 7 or more locations? What about in 2010?

In 2002, the states observed at 7 or more locations were: 
```{r seven_or_more_2002}

# Filter for 2002, determine number of locations per state, keep only those with seven or more

brfss_smart2010_overall_health %>% 
  filter(year == 2002) %>% 
  filter(response == "Excellent") %>% # Arbitarily filter by one response so there is only one observation per location
  count(state) %>%
  filter(n >= 7) %>%
  select(state) %>% 
  knitr::kable(caption = "States with seven or more locations in 2002")
```



In 2010, the states observed at 7 or more locations were: 
```{r seven_or_more seven_or_more_2010}

# Filter for 2010, determine number of locations per state, keep only those with seven or more

brfss_smart2010_overall_health %>% 
  filter(year == 2010) %>% 
  filter(response == "Excellent") %>% # Arbitarily filter by one response so there is only one observation per location
  count(state) %>%
  filter(n >= 7) %>% 
  select(state) %>% 
  knitr::kable(caption = "States with seven or more locations in 2010")
  
```


```{r q2_excellent_and_spaghetti}

# Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state.

brfss_excellent <- brfss_smart2010_overall_health %>% 
  filter(response == "Excellent") %>%
  group_by(state, year) %>% 
  summarise(excellent = mean(data_value)) 

knitr::kable(head(brfss_excellent, n=20))

#  Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years).

brfss_excellent %>% 
  ggplot(aes(x = year, y = excellent, color = state)) +
  geom_line() +
  labs(title = "Spaghetti plot of average value over time within a state")

# Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

brfss_smart2010_overall_health %>%
  filter(year %in% c(2006,2010)) %>%
  filter(state == "NY") %>%
 # filter(response %in% c("Poor", "Excellent")) %>%
  mutate(location = as.factor(locationdesc)) %>%
#  ggplot(aes(x = location, y = data_value, fill = response)) +
  ggplot(aes(x = response, y = data_value, fill = location)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~year, ncol = 1) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Distribution of data_value for responses among locations in NY in 2006 and 2010")

# Or, viewed another way, 

brfss_smart2010_overall_health %>%
  filter(year %in% c(2006,2010)) %>%
  filter(state == "NY") %>%
 # filter(response %in% c("Poor", "Excellent")) %>%
  mutate(location = as.factor(locationdesc)) %>%
  ggplot(aes(x = location, y = data_value, fill = response)) +
#  ggplot(aes(x = response, y = data_value, fill = location)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~year, ncol = 1) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Distribution of data_value for responses among locations in NY in 2006 and 2010")


```

## Problem 3

This problem has to do with accelerometers in observational studies and clinical trials. 

Load, clean, and tidy data.

```{r q3_data}
nhanes_accel <- read_csv("./data/nhanes_accel.csv")
nhanes_covar <- read_csv("./data/nhanes_covar.csv", skip = 4)

nhanes_accel <- nhanes_accel %>% 
  janitor::clean_names() %>% 
  mutate(seqn = as.character(seqn)) 
  
nhanes_covar <- nhanes_covar %>% 
  janitor::clean_names() %>% 
  filter(age > 21) %>% 
  filter(!is.na(age) & !is.na(sex) & !is.na(bmi) & !is.na(education)) %>%
  mutate(sex = as.character(sex)) %>% 
  mutate(education = as.character(education)) %>% 
  mutate(sex = case_match(sex,
                          "1" ~ "male",
                          "2" ~ "female")) %>% 
  mutate(education = case_match(education,
                                "1" ~ "less_than_high_school",
                                "2" ~ "high_school_equivalent",
                                "3" ~ "more_than_high_school")) %>%
  mutate(education = fct_relevel(education, c("less_than_high_school", "high_school_equivalent", "more_than_high_school"))) %>% 
  mutate(seqn = as.character(seqn)) %>% 
  mutate(sex = as.factor(sex))
  

# combine the two dfs

nhanes_df <- inner_join(nhanes_covar, nhanes_accel, by = "seqn")
```


Now we will explore the education distribution over demographic variables.

```{r q3_education}
# Table showing the number of men and women in each education category

education_sex <- nhanes_covar %>% 
  select(sex, education) %>% 
  group_by(education, sex) %>%
  count(education, sex) %>% 
  pivot_wider(names_from = sex,
              values_from = n)

knitr::kable(education_sex)

```

From the table, we see that the most common education level is more than high school. The distributions are relatively similar between sexes, except that there are substantially more males than females in the high school equivalent category. 

```{r q3_education_age_sex}

# visualization of the age distributions for men and women in each education category

library(ggridges)

nhanes_covar %>%
  ggplot(aes(x = education, y = age, fill = sex)) +
  geom_violin()
  geom_density_ridges(alpha = 0.2)

```


